Stage-Splitting a Modal Language

### 2014-06-20 paper/semantics.tex
More on judgements.

To understand how this behavior can be achieved, consider evaluation of the stage-\bbtwo\ term below, which contains nested $\next$ and $\prev$ blocks:
\begin{lstlisting} 
prev{
  let x = (next {1+2}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * prev{hold(#2 x)} }
}
\end{lstlisting}
Our goal is to evaluate away all of the stage-\bbone\ portions of this code, but there is more than one reasonable way to do this.
One option, taken by \cite{davies96}, is to treat $\next\{\mathtt{1+2}\}$ as a value for the purposes of stage-\bbone~evaluation, 
consequently duplicating it during substitution for {\tt x}.  
This produces:
\begin{lstlisting}
(1+2) * (1+2) * 7
\end{lstlisting}

@@

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item Partial values ($\pval$s) are stage \bbone{} terms whose stage \bbone{}
subexpressions have all been fully evaluated. In the example above,
\verb|next {x1 * x1 * 7}|.
\item Residuals ($\res$es) are stage \bbtwo{} terms whose stage \bbone{}
subexpressions have all been fully evaluated. In the example above,
\verb|1+2|.
\item Values ($\val$s) are stage \bbtwo{} terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|63|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone{} term to a pair of an
environment $\xi$ mapping newly-created variables to stage \bbtwo{} residuals,
and a partial value $v$ which may refer to those variables. This judgment
necessarily traverse stage \bbtwo{} subexpressions in order to find their stage
\bbone{} subexpressions; we call this process \emph{speculation}. Speculation is
implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo{} term to a
residual.

\TODO finish explaining how the nested evaluation actually works.
explain reification, hatted variables, and contexts.

A function $f$ which accepts inputs at both stages \bbone{} and \bbtwo{} can be
given a type of the form $A\to\fut(B\to C)$.%
\footnote{We can rewrite \texttt{fexp} in this form, or simply apply the following
higher-order function which makes the adjustment:

%\begin{lstlisting} 
%let adjust (f : $int * int -> $int) =
%	fn (p : int) => 
%		next{
%			fn (b : int) => 
%				prev{f (next {b}, p)}
%		}	
%\end{lstlisting}
\TODO code written, but listings don't work in footnotes! commented out in source
}
Given its stage \bbone{} argument $a:A$, we can evaluate the partially-applied
function:
$\cdot;\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
C)$. Next, we reify this environment into a sequence of \verb|let|-bindings
enclosing $v$, via $\reify\xi{v}{f_a}$. A canonical forms theorem on $v$
ensures that the resulting $f_a$ has type $B\to C$ in \langTwo.
Finally, given a stage \bbtwo{} argument $b:B$, we can stage-\bbtwo{} evaluate
the ultimate result of the function, $f_a~b \mathop{\tworedsym} c$.

@@

\TODO 
If we change the highlighted rules to something slightly different, then we get
precisely the semantics from \cite{davies96}. This essentially bypasses the
environment bookkeeping in $\redonesym$, by inlining residuals instead of
hoisting them in \verb|let|-bindings.

>>>>

To understand how this behavior can be achieved, consider evaluation of the stage-\bbtwo\ term below, which contains nested $\next$ and $\prev$ blocks:
\begin{lstlisting} 
prev{
  let x = (next {1+2}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * prev{hold(#2 x)} }
}
\end{lstlisting}
Our goal is to evaluate away all of the stage-\bbone\ portions of this code, but there is more than one reasonable way to do this.
One option, taken by \cite{davies96}, is to treat \verb|next{1+2}| as a value for the purposes of stage-\bbone~evaluation, 
consequently duplicating it during substitution for {\tt x}.  
This produces:
\begin{lstlisting}
(1+2) * (1+2) * 7
\end{lstlisting}

@@

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item 
Partial values ($\pval$s) are stage \bbone\ terms that have been fully evaluated, 
but which may contain hatted variables. In the example above, 
\verb|(|$\hat {\tt y}$\verb|,7)|.
\item Residuals ($\res$es) are stage \bbtwo\ terms whose stage \bbone\
subexpressions have all been fully evaluated. In the example above,
\verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
\item Values ($\val$s) are stage \bbtwo\ terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|63|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future environment} $\xi$ and a partial value $v$.
The future environment is a mapping from newly-created variables 
(which then may appear---hatted---in $v$)
to residuals.  It is the formal manifestation of the floating substitutions in our example.
This judgment, upon encountering a stage \bbtwo\ subexpression, must necessarily search into it in order to find its stage
\bbone\ subexpressions; we call this process \emph{speculation}. 
Speculation is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual.

\TODO next

\TODO prev

\TODO reification

\TODO hold

\TODO contexts

A function $f$ which accepts inputs at both stages \bbone{} and \bbtwo{} can be
given a type of the form $A\to\fut(B\to C)$%
\footnote{We can rewrite \texttt{fexp} in this form, or simply apply the following
higher-order function which makes the adjustment:

%\begin{lstlisting} 
%let adjust (f : $int * int -> $int) =
%	fn (p : int) => 
%		next{
%			fn (b : int) => 
%				prev{f (next {b}, p)}
%		}	
%\end{lstlisting}
\TODO code written, but listings don't work in footnotes! commented out in source until we get working
}.
Given its stage \bbone{} argument $a:A$, we can evaluate the partially-applied
function:
$\cdot;\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
C)$. Next, we reify this environment into a sequence of \verb|let|-bindings
enclosing $v$, via $\reify\xi{v}{f_a}$. A canonical forms theorem on $v$
ensures that the resulting $f_a$ has type $B\to C$ in \langTwo.
Finally, given a stage \bbtwo{} argument $b:B$, we can stage-\bbtwo{} evaluate
the ultimate result of the function, $f_a~b \mathop{\tworedsym} c$.

@@

\TODO 
If we change the highlighted rules to something slightly different, then we get
precisely the semantics from \cite{davies96}. This essentially bypasses the
environment bookkeeping in $\redonesym$, by inlining residuals instead of
hoisting them in \verb|let|-bindings.
From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
But moreover, because a reified residual will always be evaluated whether or not its result is consumed, 
our semantics terminates strictly less often than that of \cite{davies96}.
